---
title: 'LHR: Amsterdam Bias Analysis'
author: "Justin Braun"
date: "2024-10-13"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(openxlsx)
library(stringr)
library(friendlyeval)
```

## Load Data
```{r}
# Load confusion matrices
cms_raw <- read.xlsx('../input/Results_LHR/Output/20240308_CMs_LHR_SlimmeCheck.xlsx')
# Load feature count data
feature_counts_raw <- read.xlsx('../input/Results_LHR/Output/20240308_Important_Features_Counts.xlsx')
```

## Preprocesing

```{r}
### Confusion Matrices ###
summary(cms_raw)
print(unique(cms_raw$Dataset))
print(unique(cms_raw$Model))
print(unique(cms_raw$Feature))

cms_raw <- cms_raw %>%
  #0s indicate small sample sizes but are unlikely to be correct
  mutate(Value = ifelse(Value == 0, NA, Value))

#combine train and test since the original split is not actually recreated
cms_train <- cms_raw %>%
  filter(Dataset == 'TrainingTrain') %>%
  rename(Value_Train = Value) %>%
  dplyr::select(-Dataset)

cms_test <- cms_raw %>%
  filter(Dataset == 'TrainingTest') %>%
  rename(Value_Test = Value) %>%
  dplyr::select(-Dataset)

cms_train_test <- cms_train %>%
  left_join(cms_test, by = c('Model', 'Feature', 'Feature_Value', 'Metric')) %>%
  mutate(Value = Value_Train + Value_Test,
         Dataset = 'TrainTest') %>%
  dplyr::select(-Value_Train, -Value_Test)

cms_wide <- cms_raw %>%
  #merge combined train/test data
  filter(!(Dataset %in% c('TrainingTrain', 'TrainingTest'))) %>%
  bind_rows(cms_train_test) %>%
  
  #compute shares
  group_by(Feature, Feature_Value, Dataset, Model) %>%
  mutate(Share = (Value/sum(Value)) * 100,
         group_size = sum(Value)) %>%
  ungroup() %>%
  dplyr::select(-Value) %>%
  
  # pivot to dataset x model x group level
  pivot_wider(names_from = Metric, values_from = Share) %>%
  
  # compute various performance/fairness metrics
  mutate(TOTAL = TN+FP+TP+FN,
         ACT_N = FP + TN,
         ACT_P = FN + TP,
         PRED_P = FP + TP,
         PRED_N = FN + TN,
         FPR = (FP/ACT_N) * 100,
         PPV = (TP/PRED_P) * 100,
         FDR = (FP/PRED_P) * 100,
         TPR = (TP/ACT_P) * 100,
         STAT_PAR = PRED_P,
         ERROR = FP+FN) %>%
  
  # translate features, values, and models
  mutate(Feature_EN = case_when(Feature == 'geslacht' ~ 'gender',
                                Feature == 'Leeftijd&lt;30' ~ 'Age < 30',
                                Feature == 'Leeftijd&lt;40' ~ 'Age < 40',
                                Feature == 'Leeftijd&lt;50' ~ 'Age < 50',
                                Feature == 'IsNederlands' ~ 'Dutch',
                                Feature == 'IsWesters' ~ 'Western',
                                .default = Feature),
         Feature_Value_EN = case_when(Feature_Value == 'V' ~ 'F',
                                      Feature == 'Leeftijd&lt;30' & Feature_Value == 1 ~ 'below 30',
                                      Feature == 'Leeftijd&lt;30' & Feature_Value == 0 ~ 'above 30',
                                      Feature == 'Leeftijd&lt;40' & Feature_Value == 1 ~ 'below 40',
                                      Feature == 'Leeftijd&lt;40' & Feature_Value == 0 ~ 'above 40',
                                      Feature == 'Leeftijd&lt;50' & Feature_Value == 1 ~ 'below 50',
                                      Feature == 'Leeftijd&lt;50' & Feature_Value == 0 ~ 'above 50',
                                      Feature == 'IsNederlands' & Feature_Value == 1 ~ 'Dutch',
                                      Feature == 'IsNederlands' & Feature_Value == 0 ~ 'Not Dutch',
                                      Feature == 'IsWesters' & Feature_Value == 1 ~ 'Western',
                                      Feature == 'IsWesters' & Feature_Value == 0 ~ 'Not Western',
                                      Feature == 'IsFulltimeParent' & Feature_Value == 1 ~ 'Full-time parent',
                                      Feature == 'IsFulltimeParent' & Feature_Value == 0 ~ 'Not full-time parent',
                                      Feature == 'IsParttimeParent' & Feature_Value == 1 ~ 'Part-time parent',
                                      Feature == 'IsParttimeParent' & Feature_Value == 0 ~ 'Not part-time parent',
                                      .default = Feature_Value),
         is_privileged_group = case_when(Feature_Value == 'V' ~ 0,
                                         Feature_Value == 'M' ~ 1,
                                         Feature == 'Leeftijd&lt;30' & Feature_Value == 1 ~ 1,
                                          Feature == 'Leeftijd&lt;30' & Feature_Value == 0 ~ 0,
                                          Feature == 'Leeftijd&lt;40' & Feature_Value == 1 ~ 1,
                                          Feature == 'Leeftijd&lt;40' & Feature_Value == 0 ~ 0,
                                          Feature == 'Leeftijd&lt;50' & Feature_Value == 1 ~ 1,
                                          Feature == 'Leeftijd&lt;50' & Feature_Value == 0 ~ 0,
                                          Feature == 'IsNederlands' & Feature_Value == 1 ~ 1,
                                          Feature == 'IsNederlands' & Feature_Value == 0 ~ 0,
                                          Feature == 'IsWesters' & Feature_Value == 1 ~ 1,
                                          Feature == 'IsWesters' & Feature_Value == 0 ~ 0,
                                          Feature == 'IsFulltimeParent' & Feature_Value == 1 ~ 0,
                                          Feature == 'IsFulltimeParent' & Feature_Value == 0 ~ 1,
                                          Feature == 'IsParttimeParent' & Feature_Value == 1 ~ 0,
                                          Feature == 'IsParttimeParent' & Feature_Value == 0 ~ 1,
                                          .default = NA),
         stage = paste0(Dataset, '/', Model),
         Model_EN = case_when(Model == 'BR' ~ 'Before Reweighing',
                              Model == 'AR' ~ 'After Reweighing'))

write.csv(cms_wide, '../output/cms_wide.csv')

# convert to long
cms_long <- cms_wide %>%
  pivot_longer(cols = c("TN", "FP", "FN", "TP", "TOTAL", "ACT_N", "ACT_P", "PRED_P", "PRED_N", "FPR", "PPV", "FDR", "STAT_PAR", "TPR", "ERROR"),
               names_to = 'Metric', values_to = 'Value')

write.csv(cms_long, '../output/cms_long.csv')

# compute Amsterdam definitions
cms_amsti <- cms_wide %>%
  dplyr::select('Dataset', 'Model', 'Model_EN', 'Feature_EN', 'is_privileged_group', 'Feature_Value_EN', 'group_size', 'FP', 'FPR', 'FDR', 'PRED_P') %>%
  mutate(is_privileged_group = ifelse(is_privileged_group ==1, 'privileged', 'unprivileged')) %>%
  pivot_wider(names_from = 'is_privileged_group', values_from =  c('Feature_Value_EN', 'group_size', 'FP', 'FPR', 'FDR', 'PRED_P')) %>%
  mutate(FP_amsti_diff = 100*((FP_unprivileged-FP_privileged)/FP_privileged),
         FPR_amsti_diff = 100*((FPR_unprivileged-FPR_privileged)/FPR_privileged),
         FDR_amsti_diff = 100*((FDR_unprivileged-FPR_privileged)/FDR_privileged),
         PRED_P_amsti_diff = 100*((PRED_P_unprivileged-PRED_P_privileged)/PRED_P_privileged))
write.csv(cms_amsti, '../output/cms_amsti.csv')
```

```{r}
### Feature Importance ###
feature_counts <- feature_counts_raw %>%
  mutate(Count = ifelse(Count == 0, NA, Count),
         # translate
         Feature_EN = case_when(Feature == 'geslacht' ~ 'gender',
                                Feature == 'Leeftijd&lt;30' ~ 'Age < 30',
                                Feature == 'Leeftijd&lt;40' ~ 'Age < 40',
                                Feature == 'Leeftijd&lt;50' ~ 'Age < 50',
                                Feature == 'IsNederlands' ~ 'Dutch',
                                Feature == 'IsWesters' ~ 'Western',
                                .default = Feature),
         Feature_Value_EN = case_when(Feature == 'geslacht' & Value == 1 ~ 'F', #not sure about gender here
                                      Feature == 'geslacht' & Value == 0 ~ 'M',
                                      Feature == 'Leeftijd&lt;30' & Value == 1 ~ 'below 30',
                                      Feature == 'Leeftijd&lt;30' & Value == 0 ~ 'above 30',
                                      Feature == 'Leeftijd&lt;40' & Value == 1 ~ 'below 40',
                                      Feature == 'Leeftijd&lt;40' & Value == 0 ~ 'above 40',
                                      Feature == 'Leeftijd&lt;50' & Value == 1 ~ 'below 50',
                                      Feature == 'Leeftijd&lt;50' & Value == 0 ~ 'above 50',
                                      Feature == 'IsNederlands' & Value == 1 ~ 'Dutch',
                                      Feature == 'IsNederlands' & Value == 0 ~ 'Not Dutch',
                                      Feature == 'IsWesters' & Value == 1 ~ 'Western',
                                      Feature == 'IsWesters' & Value == 0 ~ 'Not Western',
                                      Feature == 'IsFulltimeParent' & Value == 1 ~ 'Full-time parent',
                                      Feature == 'IsFulltimeParent' & Value == 0 ~ 'Not full-time parent',
                                      Feature == 'IsParttimeParent' & Value == 1 ~ 'Part-time parent',
                                      Feature == 'IsParttimeParent' & Value == 0 ~ 'Not part-time parent',
                                      .default = as.character(Value))) %>%
  group_by(Feature_EN, Feature_Value_EN, dataset) %>%
  mutate(share = (Count/sum(Count, na.rm = T)) * 100) %>% #note that I remove NAs which are presumably not in fact all zeros but could be low values below 10
  ungroup()
```

## RQ 1 How did Reweighing affect various fairness metrics?

```{r}
# subset to relevant definitions, datasets, and characteristics
cms_city_perspective <- cms_long %>%
  filter(stage %in% c('Prepilot/BR', 'Prepilot/AR'), 
         Metric %in% c('STAT_PAR', 'FDR', 'FP'),
         Feature_EN %in% c('gender', 'Age < 30', 'Age < 50', 'Dutch', 'IsFulltimeParent', 'Western')) %>%
  mutate(order = case_when(stage == 'Prepilot/BR' ~ 1,
                           stage == 'Prepilot/AR' ~ 2,
                           .default = NA))

cms_char_diff <- data.frame()

#loope over characteristics
for(characteristic in c('Dutch', 'gender', 'IsFulltimeParent')){
  cms_char <- cms_city_perspective %>%
    filter(Feature_EN == characteristic)
  
  # compare metrics by group
  p1 <- ggplot(cms_char, aes(x = reorder(Model_EN, order), y = Value, fill = Feature_Value_EN))+
    geom_bar(stat = 'identity', position = position_dodge())+
    facet_grid(Metric ~., scales = "free_y")+
    labs(x = 'Dataset/Model',
         title = paste0(characteristic, ': Development of Fairness Metrics from the Citys Perspective'),
         fill = characteristic)+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

  print(p1)
  ggsave(paste0('../output/rq1_p1_', characteristic, '.png'), plot = p1, width = 10, height = 8)
  
  # compute Amsterdam-style comparisons
  feature_val_privileged <- unique(cms_char[cms_char$is_privileged_group == 1,]$Feature_Value_EN)[1]
  feature_val_unprivileged <- unique(cms_char[cms_char$is_privileged_group == 0,]$Feature_Value_EN)[1]

  cms_char_diff <- cms_char %>%
    dplyr::select(-Feature_Value, -group_size, -is_privileged_group) %>%
    pivot_wider(names_from = 'Feature_Value_EN', values_from = 'Value') %>%
    #using Amsterdam's difference op here, though not sure the ref cat is always the same
    mutate(Diff = 100*(.data[[feature_val_unprivileged]] -  .data[[feature_val_privileged]])/.data[[feature_val_privileged]],
           Metric = case_when(Metric == 'FP' ~ 'False Positive Share',
                              Metric == 'FDR' ~ 'False Discovery Rate',
                              Metric == 'STAT_PAR' ~ 'Statistical Parity',
                              .default = NA))

  # plot difference as operationalized by the city reports
  p2 <- ggplot(cms_char_diff, aes(x = reorder(Model_EN, -order), y = Diff))+
    geom_bar(stat = 'identity', position = position_dodge(), fill = 'blue') +
    facet_grid(Metric ~., scales = "free_y")+
    labs(x = 'Dataset/Model',
         y = 'Group % difference',
         title = paste0(characteristic, ': Development of Fairness \nMetrics from the Citys Perspective'),
         subtitle = paste0(feature_val_unprivileged, ' - ', feature_val_privileged))+
    coord_flip()

  print(p2)
  ggsave(paste0('../output/rq1_p2_', characteristic, '.png'), plot = p2, width = 10, height = 8)
  
  write.csv(cms_char_diff, paste0('../output/fairness_definition_cmomp_', characteristic, '.csv'))
}
```



## RQ2 Error rate in Pilot was much higher than in tests

```{r}
# subset to relevant characteristics and metrics
cms_error <- cms_long %>%
  filter(stage %in% c( 'Prepilot/AR', 'Pilot/AR'),
         Metric %in% c('FP', 'FN', 'ERROR'),
         Feature_EN == 'Dutch') %>%
  mutate(order = case_when(stage == 'Prepilot/AR' ~ 1,
                           stage == 'Pilot/AR' ~ 1,
                           .default = NA))

# produce plots for error rate for Dutch vs non-Dutch applicants
p4 <- ggplot(cms_error, aes(x = reorder(stage, order), y = Value, fill = Metric))+
  geom_bar(stat = 'identity', position = position_dodge())+
  facet_wrap(.~Feature_Value_EN)+
 labs(x = 'Dataset/Model', y = '% Error',
       title = paste0(characteristic, ': Development of Error rates across model development'),
       fill = 'Error Metric')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
print(p4)

ggsave(paste0('../output/rq3_p4_error_', characteristic, '.png'), plot = p4, width = 10, height = 8)

cms_error <- cms_error %>%
  mutate(Metric = case_when(Metric == 'FP' ~ 'False Positives',
                            Metric == 'FN' ~ 'False Negatives',
                            Metric == 'ERROR' ~ 'Overall Error Rate'))
write.csv(cms_char, paste0('../output/', characteristic ,'_error_rate.csv'))


```

## RQ 3 Feature importance

Along with the classification, the model provided caseworkers with the
three most important features used by the model to come to its
determination. Loek provided us access to the most important feature by
demographic group. This allows us 1) to see if caseworkers could deduce
beneficiary charactsitics from the highlighted features, potentially
activating their biases, and 2) whether the model used different
features for different demographic groups in coming to its
determination. The latter could be concerning under due process
considerations.

```{r}
feature_counts_restricted <- feature_counts %>%
  filter(Feature_EN %in% c("gender", "IsFulltimeParent", "Dutch")) %>%
  group_by(Feature_EN, Feature_Value_EN, dataset) %>%
  arrange(desc(share)) %>%
  mutate(rank = dense_rank(desc(share))) %>%
  slice_max(n = 5, order_by = share) %>%
  ungroup() %>%
  filter(!is.na(share))

# plot top 5 most important feature by applicant characteristic and development stage
for(characteristic in unique(feature_counts_restricted$Feature_EN)){
  feature_counts_char <- feature_counts_restricted %>%
    filter(Feature_EN == characteristic)
  
  p3 <- ggplot(feature_counts_char, aes(x = rank, y = share, fill = Feature_Value_EN, label = Important.Feature))+
    geom_bar(stat = 'identity', position = position_dodge())+
    geom_text(aes(y = 0), hjust = 0, angle = 90, position = position_dodge(width = .9))+
    facet_grid(.~dataset)+
    labs(x = 'importance rank',
         y = '% top 3 most important features',
         title = paste0(characteristic, ': Most important Features'))
  
  print(p3)
  ggsave(paste0('../output/rq3_feature_importance_', characteristic, '.png'), plot = p3, width = 10, height = 8)
  
}
```



